The Journey from Wild to Text Book Data: A Case Study of the National Longitudinal Survey of Youth

### 2021-08-12 Report/report_alt.Rmd
Calculate gender wage gap and fit linear mixed model.

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE,
                      cache = TRUE, 
                      cache.path = "cache/",
                      fig.path = "figures/")
library(tidyverse)
library(brolgar)
library(patchwork)
library(kableExtra)
library(MASS)
library(janitor)
library(DiagrammeR)
library(rsvg)
library(webshot)
library(mgcv)
library(modelr)
library(tsibble)
```

@@

```{r rlm, cache=TRUE, warning = FALSE, message = FALSE, echo = FALSE}

@@

# if the weight < 1, the mean_hourly_wage is replaced by the model's fitted/predicted value.
# and add the flag whether the observation is predicted value or not.
# since the fitted value is sometimes <0, and wages value could never be negative,
# we keep the mean hourly wage value even its weight < 1.

@@

@@

sample_id <- sample(unique(wages_demog_hs$id), 20)
sample <- subset(wages_demog_hs, id %in% sample_id)

@@

In the second example, we examined the pattern of mean hourly wage corresponding to different groups of gender and race. We took 240 samples of females and males equally and spread them into twelve facets, as shown in Figure \ref{fig:plot-gender} using `facet_strata` function from `brolgar` [@brolgar]. We learn that the mean hourly wage of females and males fluctuated over the years. For example,  in facets 4, 5, and 10, males tend to earn more wages than females. This finding is also shown in Figure \ref{fig:se-gender} where we plotted the median wage of males and females against the year. We learn that males had a higher mean of mean hourly wage throughout the survey than females. Unfortunately, the gap between them tends to be wider from time to time.

@@

```{r se-gender, echo=FALSE, fig.cap="The yearly mean of mean hourly wage by gender along with its standard error. Males generally earn more wages than females. The error interval tend to be wider over the years."}
wages_hs2020_grouped_gender <- wages_hs2020 %>%
  group_by(gender, year) %>%
  summarise(mean = mean(mean_hourly_wage),
             sd = sd(mean_hourly_wage),
             samplesize = n()) %>%
  ungroup() %>%
  mutate(upper = mean + sd/sqrt(samplesize),
         lower = mean - sd/sqrt(samplesize))

gender_plot <- ggplot(wages_hs2020_grouped_gender,
                      aes(x = year,
                          y = mean,
                          group = gender,
                          colour = gender)) +
  geom_point(size = 1) +
  geom_errorbar(aes(ymin=lower, ymax=upper), width=1,
                 position=position_dodge(0.1)) +
  theme(legend.position = "bottom",
         legend.text = element_text(size = 5),
         axis.title = element_text(size = 9)) +
  scale_color_brewer(palette = "Dark2") +
  ylab("median of mean hourly wage (US$)")

gender_plot

For the final example, we showed modeling the longitudinal data using the Generalized Additive Model (GAM) with `mgcv` package [@mgcv] for each ID. We fitted a more flexible model to fit into the data since the relationship between mean hourly wage and year is not linear due to the spikes. Since this approach is computationally expensive, we only showed the model in a small fraction of the data. We sampled 1 percent of the total respondents using `brolgar's` `sample_frac_keys` function. Further, we plotted some IDs overlayed by the fitted GAM model in Figure \ref{fig:plot-gam}. We learn that the model flexible enough to deal with the fluctuation in the data. Moreover, the data with small spikes tend to be linearly fitted.

@@

# take a small fraction of key
set.seed(2020)
wages_sample_gam <- sample_frac_keys(wages_hs2020_tsibble, size = 0.01) %>%
  mutate(id_fct = as.factor(id),
         year0 = year - 1979)

# build GAM model
wages_gam <- gam(
  mean_hourly_wage ~ s(year0, by = id_fct) + id_fct,
  data = wages_sample_gam,
  method = "REML"
)

wages_aug_gam <- wages_sample_gam %>%
  add_predictions(wages_gam, var = "pred") %>%
  add_residuals(wages_gam, var = "res") %>%
  group_by_key() %>%
  mutate(rss = sum(res^2)) %>%
  ungroup()

set.seed(2021)
wages_aug_gam %>%
  sample_n_keys(12) %>%
  ggplot(aes(x = year,
             y = pred,
             group = id)) +
  geom_line(color = "steelblue") +
  geom_point(aes(y = mean_hourly_wage)) +
  facet_wrap(~id)

>>>>

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE,
                      #cache = TRUE, 
                      #cache.path = "cache/",
                      fig.path = "figures/")
library(tidyverse)
library(brolgar)
library(patchwork)
library(kableExtra)
library(MASS)
library(janitor)
library(DiagrammeR)
library(rsvg)
library(webshot)
library(mgcv)
library(modelr)
library(tsibble)
library(lme4)
library(modelr)
```

@@

```{r rlm, eval=FALSE, warning = FALSE, message = FALSE, echo = FALSE}

@@

# if the weight < 0.12, the mean_hourly_wage is replaced by the model's fitted/predicted value.
# and add the flag whether the observation is predicted value or not.
# since the fitted value is sometimes <0, and wages value could never be negative,
# we keep the mean hourly wage value even its weight < 0.12.

@@

# save it to rds file so it would faster the knitting process
saveRDS(wages_demog_hs, here::here("Report/Result/wages_demog_hs_tmp.rds"))
```

@@

wages_demog_hs <- readRDS(here::here("Report/Result/wages_demog_hs_tmp.rds"))
sample_id <- sample(unique(wages_demog_hs$id), 20)
sample <- subset(wages_demog_hs, id %in% sample_id)

@@

In the second example, we examined the pattern of mean hourly wage corresponding to different groups of gender and race. We took 240 samples of females and males equally and spread them into twelve facets, as shown in Figure \ref{fig:plot-gender} using `facet_strata` function from `brolgar` [@brolgar]. We learn that the mean hourly wage of females and males fluctuated over the years. For example,  in facets 4, 5, and 10, males tend to earn more wages than females.

@@

To further investigate this, we calculated the median wage of females relative to males, which is called gender wage gap. According to @OECD, gender pay gap is the difference between median wage of males and females relative to median wage of males. We find that, as shown in \ref{fig:se-gender}, over the years, females tend to earn fewer than males. For example, in 2018 females earn about 78 cents for every $1 earned by males. The widest gender gaps were observed in 1990s.

```{r se-gender, echo=FALSE, fig.cap="Gender wage gap or the median wages of females relatives to median wages of males. Females tend to earn fewer than males. For example, in 2018 females earn about 78 cents for every $1 earned by males."}

wages_hs2020_grouped_gender <- wages_hs2020 %>%
  group_by(gender, year) %>%
  summarise(mean = mean(mean_hourly_wage),
            median = median(mean_hourly_wage),
             sd = sd(mean_hourly_wage),
             samplesize = n()) %>%
  ungroup() %>%
  mutate(upper = mean + sd/sqrt(samplesize),
         lower = mean - sd/sqrt(samplesize))

wage_female <- wages_hs2020_grouped_gender %>%
  filter(gender == "FEMALE") %>%
  dplyr::select(year, median) %>%
  rename(female = median)

wage_male <- wages_hs2020_grouped_gender %>%
  filter(gender == "MALE") %>%
  dplyr::select(median) %>%
  rename(male = median)

gender_wage_gap <- cbind(wage_female, wage_male) %>%
  mutate(gap = (male-female)/male) %>%
  mutate(relative = 1-gap)

ggplot(gender_wage_gap) +
  geom_line(aes(x = year,
                y = relative),
            color = "#DF7401", size = 0.8) +
  geom_point(aes(x = year,
                y = relative),
             color = "#DF7401", size = 2) +
  geom_hline(yintercept = 1, colour = "#088A68", size = 1) +
  scale_x_continuous(limits = c(1978.7,2018.3), expand = c(0, 0)) +
  ylab("females' median wages relative to males (US$)") +
  annotate("text", x = 2014.5, y = 1, label = "$1 of male wages", vjust = 1.4) +
  annotate("text", x = 2012, y = 0.8, label = "female wages relative to male", vjust = 1)
```

For the final example, we showed modeling the longitudinal data using linear mixed model with `lme4` package [@lme4].

We fitted a more flexible model to fit into the data since the relationship between mean hourly wage and year is not linear due to the spikes. Since this approach is computationally expensive, we only showed the model in a small fraction of the data. We sampled 1 percent of the total respondents using `brolgar's` `sample_frac_keys` function. Further, we plotted some IDs overlayed by the fitted GAM model in Figure \ref{fig:plot-gam}. We learn that the model is flexible enough to deal with the fluctuation in the data. Moreover, the data with small spikes tend to be linearly fitted.

@@

# transform the data type and year for modeling
wages_model <- wages_hs2020_tsibble %>%
  mutate(id_fct = as.factor(id),
         year0 = year - 1979,
         ln_wages = log(mean_hourly_wage))

# build GAM model
wages_fit <- lmer(
  ln_wages ~ year0 + hgc_i + gender + (1|id),
  data = wages_model
)

wages_fit_aug<- wages_model %>%
  add_predictions(wages_fit, var = "pred") %>%
  add_residuals(wages_fit, var = "resid")

slope_only <- wages_slope %>%
  dplyr::select(id, .slope_year) %>%
  arrange(desc(.slope_year)) %>%
  slice_head(n = 12)

wages_fit_aug2 <- wages_fit_aug %>%
  filter(id %in% slope_only$id)

ggplot(wages_fit_aug2, aes(x = year,
             y = pred,
             group = id,
             colour = factor(id))) + 
  geom_line() + 
  geom_point(aes(x = year,
                 y = ln_wages,
                 colour = factor(id))) + 
  facet_wrap(~id) + 
  theme(legend.position = "none")
